---
title: "A Simple Ratio for Estimating the Effective Reproduction Value of an Epidemic"
authors:
- name: Authors
  email: emails
  address: institutions
abstract: |
  ABSTRACT
output:
  bookdown::word_document2:
    number_sections: no
    extra_dependencies: float
    keep_tex: yes
    includes:
      in_header: header.tex
  bookdown::pdf_book:
    base_format: rticles::oup_article
    number_sections: no
    extra_dependencies: float
    keep_tex: yes
    includes:
      in_header: header.tex
header-includes:
  - \usepackage[labelsep=period,labelfont=bf]{caption}
  - \usepackage{hyperref}
  - \usepackage[capitalize, nameinlink]{cleveref}
  - \crefdefaultlabelformat{#2\textbf{#1}#3} % <-- Only #1 in \textbf
  - \crefname{figure}{\textbf{Figure}}{\textbf{Figures}}
  - \Crefname{figure}{\textbf{Figure}}{\textbf{Figures}}
  - \crefname{table}{\textbf{Table}}{\textbf{Tables}}
  - \Crefname{table}{\textbf{Table}}{\textbf{Tables}}
bibliography:
  - packages.bib
  - references.bib
keywords:
  - key
  - dictionary
  - word
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE) # By default, hide code; set to TRUE to see code
knitr::opts_chunk$set(fig.pos = 'p') # Places figures on pages separate from text
knitr::opts_chunk$set(out.width = '100%', dpi=300) # Figure resolution and size
knitr::opts_chunk$set(fig.env="figure") # Latex figure environment
knitr::opts_chunk$set(eval.after = "fig.cap")
knitr::opts_chunk$set(echo = F,
                      message = F, 
                      warning = F)

## Load packages
library("tidyverse")
library("ggplot2")
library("ggpubr")
library("drake")
library("english")
library("bibtex")
library("here")

# Set working directory
here::i_am("docs/manuscript/manuscript.Rmd")

# Get Figure creation functions
source(here::here("code", "create_figures.R"))

        

# Convert number to text 
# Ref: https://5harad.com/mse125/r/visualization_code.html
addUnits <- function(n) {
  labels <- ifelse(n < 1000, n,  # less than thousands
                   ifelse(
                     n < 1e6,
                     paste0(round(n / 1e3), ' thousand'),
                     # in thousands
                     ifelse(
                       n < 1e9,
                       paste0(round(n / 1e6), ' million'),
                       # in millions
                       ifelse(n < 1e12, paste0(round(n / 1e9), ' billion'), # in billions
                              'too big!')
                     )
                   ))
  return(labels)
}

# Covert a vector to a english list
vector_to_englishList <- function(x) {
  last_item = tail(x, 1)
  
  englishList = paste(head(x,-1), collapse = ", ")
  englishList = paste(englishList, last_item, sep = ", and ")
  
  return(englishList)
}

# Function between to numbers
```

```{r load-data}
simulation_parameters = readd("simulation_parameters")
rt_estimates = readd("rt_estimates")
national_data = readd("national_data")
COVID19_params = readd("COVID19_params")

# Comparison with COVID-19 data
national_data_filtered = national_data %>% 
  filter(time > as.Date("2020-04-01") & time < as.Date("2020-12-01")) %>% 
  filter(cases_new > 19)


estimation_diff = national_data_filtered$estimation_diff
estimation_diff_median = estimation_diff %>%  median(., na.rm=T) %>% signif(digits = 2)
estimation_diff_quantile_975 = estimation_diff %>% quantile(., 0.975, na.rm = T) %>% signif(digits = 2) %>% format(scientific = F)
estimation_diff_quantile_025 = estimation_diff %>% quantile(., 0.025, na.rm = T) %>% signif(digits = 2) %>% format(scientific = F)

# Get packages used for this project
packages <- readLines(here::here("code", "packages.R")) %>% #read file
              sub('[^\"]+\"([^\"]+).*', '\\1', .)    
package_citation_ids =  read.bib(here("./docs/manuscript", "packages.bib")) %>% 
  names(.) %>% regmatches(., m = regexpr(pattern = "^R-.+", text = .)) %>% 
  paste("@", ., sep="") %>% paste(collapse = "; ")
```

# Introduction {.unnumbered}

The effective reproduction number ($R_t$) is a fundamental measure in infectious disease epidemiology. It can be defined as the average number of secondary cases caused by an infected individual. The $R_t$ changes over the course of an epidemic, because it is not only determined by characteristics of the pathogen but also by human behavior (e.g. face mask usage, physical distancing, and travel restrictions). While the $R_t$ is a retrospective measure and does not predict future behavior of an epidemic, observing the $R_t$ consistently above 1, indicates that the epidemic is growing exponentially. If, by contrast, the $R_t$ is consistently below 1, it indicates that epidemic is shrinking. Furthermore, the $R_t$ can vary between geographic locations, as social behaviors and environmental factors vary between these locations.

As the COVID-19 pandemic has demonstrated, communication to the public and political leadership regarding infection rates is critical to management of the pandemic. The $R_t$ can be used to communicate whether an epidemic is growing or shrinking in a region. However, this communication may be hindered by the fact that many of the current methods for estimating $R_t$ require advanced mathematics to understand and to apply [@cori_new_2013; @wallinga_different_2004; @wallinga_how_2007].

Here, we describe how $R_t$ can be approximated by a simple ratio of new infections on day $t$ and day $t-d$, where $d$ is the mean generation interval of the infection (\cref{eq:simple-ratio-eq}). The *generational interval* being the time between when a primary case infected to the when they infect a secondary case.

Our aim is two-fold: First, by describing a simplified method of calculating $R_t$, we aim to allow others to easily and rapidly calculate $R_t$. Second, we aim to assess the accuracy of the simple ratio compared to the more advanced methods for estimating the $R_t$. This second aim is important, because this simplified ratio has been used by public health bodies to report the $R_t$ during the COVID-19 pandemic - e.g. reports created by Germany's Robert Koch Institute [@Heiden2020Schatzung] - but we have not found significant literature describing the performance of the method (at least in published English literature).

```{=tex}
\begin{equation}
  R_t \approx \frac{I_t}{I_{t-d}}
  (\#eq:simple-ratio-eq)
\end{equation}
```
## Deriving the Simple Ratio for estimating $R_t$ {.unnumbered}

Among the previously developed methods estimating $R_t$, it common to base the estimation on the following relationship:

```{=tex}
\begin{equation} 
  R_t = \frac{I_t}{\sum_{a=0}^\infty{I_{t-a}g(a)}}
  (\#eq:original-rt)
\end{equation}
```
$I_t$ is the number of new cases produced on day $t$; $g(a)$ is the probability density function describing the probability that a person infected $a$ days ago will cause a new case today (at time $t$) - $g(a)$ can be thought of as equivalent to to the generation interval distribution. Intuitively, this equations can be understood in the following way. Consider a scenario where for every day prior to today, only one person was infected. In such a scenario, the denominator of \cref{eq:original-rt} would equal 1 (since it would be the sum of all values of probability density function) and the $R_t$ would equal $I_t$. In this framework $R_t$ is equal to the number of new infections created today by a cohort of people that span the entire lifetime of an infection, which is mathematically equivalent to the traditional framework of $R_t$ as the total number of new cases generated by one infected person across the entire lifetime of that person's infection. Now consider the scenario where multiple people were infected each day prior to today. In such a scenario, each additional infected person would contribute an additional probability of new cases and increase the value of $I_t$; therefore to estimate $R_t$, $I_t$ must be scaled down by the additional probability each previously infected person contributes - hence the denominator in \cref{eq:original-rt}. To understand the mathematical derivation of this equation, we refer readers to the article by Wallinga and Lipsitch [-@wallinga_how_2007].

If we make the simplifying assumption that there is minimal variance in $g(a)$ (i.e. $g(a)$ is 1 when $a$ equals the mean generation interval ($d$) and $g(a)$ is zero for all other values of $a$), then Equation \cref{eq:original-rt} can be simplified to:

```{=tex}
\begin{equation}
  R_t \approx \frac{I_t}{I_{t-d}}
  (\#eq:simple-ratio-eq)
\end{equation}
```
We can make this estimation of $R_t$ robust to variations in the generation interval and reporting of new cases by calculating $R_t$ as the ratio of the sums of new cases within two time windows size $\tau$ days (with the center of the time windows being on days $t$ and $t-d$):

```{=tex}
\begin{equation} 
  R_t \approx \frac{\sum_{t - \frac{\tau}{2}}^{t + \frac{\tau}{2}}I_i}{\sum_{t - \frac{\tau}{2}}^{t + \frac{\tau}{2}}I_{i-d}} = \frac{n_{t}}{n_{t-d}}
  (\#eq:eq6)
\end{equation}
```
A confidence interval can be generated for this ratio by carrying forward a confidence interval of the mean generation interval.

# Materials and methods {.unnumbered}

## Simulation {.unnumbered}

A deterministic Susceptible-Exposed-Infectious-Resolved (SEIR) model was used to create epidemic simulations. No birth rate, death rate, or R to S transition rates were included. We included `r simulation_parameters$Nsims` simulated epidemics in our analysis. For each epidemic, the mean generation interval of the infection was randomly selected from `r simulation_parameters$mean_gi_range[1]` to `r simulation_parameters$mean_gi_range[2]` days. The generation interval was randomly divided into latent and infectious periods. Each simulation was divided in to `r as.english(simulation_parameters$num_periods[1])` `r simulation_parameters$period_lengths[1]` day periods with differing $R_0$ values (representing changes in social behaviors in the populations); and the $R_0$ in these periods was randomly selected from `r simulation_parameters$r0_range[1]` to `r simulation_parameters$r0_range[2]`. Between each `r simulation_parameters$period_lengths[1]` day period, a transition period was included in which the $R_0$ would gradually change following a logistic curve. The length of the transition period was randomly selected from `r simulation_parameters$transition_lengths[1]` to `r simulation_parameters$transition_lengths[2]` days. Only simulations that lasted at least `r simulation_parameters$min_simLength` days and never had less than `r simulation_parameters$min_newCases` cases per day were included in final analysis.

## $R_t$ Estimation {.unnumbered}

Custom R code was used to generate the simple ratio $R_t$ estimations, and the EpiEstim R package was used to generate the Cori method $R_t$ estimations. See below for information on code availability. When assessing the accuracy of both methods, estimations made prior to the 15th day of each simulation was not included because both methods, as implemented here, are significantly less accurate during this early period of the epidemic.

## COVID-19 Data {.unnumbered}

Data on daily COVID-19 cases were acquired from the covidregionaldata R package [@R-covidregionaldata]

## Code and Data Availability {.unnumbered}

All simulations, data processing, analysis, and graphics were all created using the R statistical software using the following packages: `r vector_to_englishList(packages)` [`r package_citation_ids`]. All code and data used to develop this project is available at <https://github.com/aryazand/simple_rt>

# Results {.unnumbered}

## Accuracy of the simple ratio estimation of $R_t$ {.unnumbered}

```{r}
# Descriptive Statistics
N = simulation_parameters$initial_state_values %>% sum()
error_summary = rt_estimates %>% filter(time > 49) %>% 
  select(matches("rt.+_mean_error")) %>% 
  rename_with(~ gsub("rt\\.(.+)_mean_error", "\\1", .x)) %>%
  transmute_all(abs) %>% 
  summarise_all(quantile, probs = c(0.025, 0.5, 0.975), na.rm = T) %>% 
  transmute_all(signif, digits = 2)
```

We assessed how accurately the simple ratio can estimate $R_t$ compared to a standard method developed by Cori et al [-@cori_new_2013] - which we will refer to as the "Cori method". $R_t$ calculation with either the simple ratio and Cori method requires knowing the mean generation interval of the infection; however, the Cori method also requires knowing the standard deviation of the generation interval or having an empiric distribution of the generation interval.

To assess accuracy of $R_t$ estimations, we used Susceptible-Exposed-Infectious-Resolved (SEIR) model to simulate epidemics and then compare how accurately the two methods can estimate the $R_t$ values in the epidemics. We simulated `r simulation_parameters$Nsims` epidemics in a population of `r addUnits(N)` individuals. We varied the lengths of latent and infectiousness periods with each simulation. In each simulation, the population underwent changes in the $R_t$ value representing changes in population behavior (e.g. "lockdown" period or lifting of a lockdown). \cref{fig:sim-results}**A** shows the true $R_t$ and the estimated $R_t$ values for each day in one such simulated epidemic. A full visualization of all the simulations can be seen in (\cref{fig:all-simulations}).

Over the `r simulation_parameters$Nsims` simulation, the median estimation error of the daily $R_t$ by the simple ratio method was `r error_summary$simple['50%']` (95% inter-percentile range of `r error_summary$simple['2.5%']` to `r error_summary$simple['97.5%']`) (\cref{fig:sim-results}**B**). By comparison, the median estimation error of the Cori method was `r error_summary$cori['50%']` (95% inter-percentile range of `r error_summary$cori['2.5%']` to `r error_summary$cori['97.5%']`).

We identified two factors contributing to error by the simple ratio. First, the simple ratio has a tendency to over-estimate the $R_t$, especially as the true value of $R_t$ increases above 1.5 (\cref{fig:sim-results}**C**). This is consistent with previous mathematical analysis of this simplified method of estimating $R_t$ [@wallinga_how_2007]. Second, the simple method (like the Cori method) is more prone to error when there is a rapid change in $R_t$ - i.e. going in and out of lockdown (\cref{fig:sim-results}**D**). When $R_t$ is relatively stable, the simple method may over-estimate $R_t$; however during periods of rapid change in $R_t$ (e.g. going in and out of lockdown), the simple method may either over- or under-estimate $R_t$. Overall, while the Cori method performs better, the simple method provides an estimation within a reasonable margin of error.

(ref:sim-results-cap) \textbf{The Simple Ratio approximates the effective reproduction in simualted epidemics.} One hundred epidemics were simulated using an SEIR model and the simple ratio and Cori method were used to estimate the known $R_t$ over time for each simulated epidemic. \textbf{(A)} The true $R_t$ (black line) and estimated $R_t$ (colored lines) over time for one of simulated epidemics. \textbf{(B)} The distribution of the absolute errors of both estimation methods across all simulated epidemics. The median absolute error for the Cori and simple method were `r error_summary$cori['50%']` (95% IPR `r error_summary$cori['2.5%']` to `r error_summary$cori['97.5%']`) and `r error_summary$simple['50%']` (95% IPR `r error_summary$simple['2.5%']` to `r error_summary$simple['97.5%']`), respectively. Note only error after the 50th day was considered because error early in the pandemic was inordinately large for both methods, as shown in Panel D. \textbf{(C)} The estimation error of the simple ratio method is associated with size of true $R_t$. \textbf{(D)} The estimation error of both methods over time for the simulated epidemics. Each line represents the error in $R_t$ estimation for a given simulation. Note that for both methods, a substantial proportion of error occurs around days 60 and 130 when there was is rapid change in $R_t$. IPR = "inter-percentile range".

```{r sim-results, fig.cap = '(ref:sim-results-cap)'}

# Plots
panel_A <- rt_estimates %>%
  filter(simulation_run == 66) %>%
  create_estimate_plot()

panel_B <- rt_estimates %>%
  filter(time > 49) %>%
  create_rt_estimation_error_boxplot()

panel_C <- rt_estimates %>%
  filter(time > 49) %>%
  create_rt_estimation_error_by_rt()

panel_D <- rt_estimates %>% 
  pivot_longer(cols = ends_with("mean_error"), names_to = "Method", values_to = "estimation_error") %>% 
  mutate(Method = recode(Method, rt.cori_mean_error = "cori", rt.simple_mean_error = "simple")) %>%
  ggplot(aes(x = time, y = estimation_error)) + geom_line(aes(group = interaction(simulation_run, Method), color = Method), alpha = 0.2) + 
  facet_wrap(~Method, ncol = 2) +
  ylab("Estimation Error") + 
  custom_theme() + 
  theme(strip.background = element_blank(), 
        strip.text = element_blank(),
        legend.position = "none") + 
  scale_color_manual(values=cbPalette) +
  scale_fill_manual(values=cbPalette) + 
  ylim(-0.25, 0.5)

full_figure_top = ggarrange(panel_A, panel_B, panel_C, ncol = 3, align = "hv", labels=c("A", "B", "C"), common.legend = T)
full_figure_bottom = ggarrange(panel_D, labels=c("D"))
ggarrange(full_figure_top, full_figure_bottom, nrow = 2, common.legend = T)

```

Often the the mean and standard deviation of the generation interval is not known exactly. For instance, for the COVID-19 the mean generation interval estimates, with 95% credible intervals, include 3.95 (3.01 - 4.91) and 5.20 (3.78 - 6.78); and estimates of the standard deviation of the generation interval include 1.51 (0.74 - 2.97) and 1.72 (0.91 - 3.93) [@ganyani_estimating_2020]. Therefore, we compared the accuracy of the two methods if the standard deviation of the generation interval is mis-specified as three days longer. Under these conditions, the simple ratio actually performs almost identically to the Cori method (\cref{fig:estimation-inaccParams}).

(ref:real-world-data-cap) \textbf{The Simple Ratio and Cori Method estimate similar Rt values for the COVID-19 pandemic}. \textbf{(A)} Rt estimations for four sample nations \textbf{(B)} Violin plot displaying the absolute difference in Rt estimation across for each day of the pandemic (Apr 01 to Dec 12) for all 233 individual nations. Median absolute difference in estimations was `r estimation_diff_median` (95% IPR `r estimation_diff_quantile_025` to `r estimation_diff_quantile_975`). IPR = "inter-percentile range".

```{r real-world-data, fig.cap = "(ref:real-world-data-cap)"}

countries = c("United States", "Canada", "United Kingdom", "Germany")

panel_A <- national_data %>%
  filter(country %in% countries) %>%
  filter(time %in% seq(as.Date("2020-10-01"), as.Date("2020-10-31"), by = "day")) %>%
  pivot_longer(., cols = starts_with("rt."), names_to = c("Method", "parameter"), 
              names_pattern = "rt.([[:alpha:]]+)_([[:print:]]+)", values_to = "value") %>%
  pivot_wider(names_from = "parameter", values_from = "value") %>%
  ggplot() + 
    geom_line(aes(time, mean, group = Method, color = Method), size = 1) +
    geom_ribbon(aes(time, ymin = quantile_025, ymax = quantile_975, group = Method, fill = Method), alpha = 0.2) + 
    geom_hline(yintercept = 1, linetype = 2) + 
    facet_wrap(~country, scales = "free") + 
    ylab("Rt") + 
    custom_theme() + 
    theme(strip.background = element_blank(),
          legend.position = "top") +
    scale_color_manual(values=cbPalette)

panel_B <- national_data_filtered %>%
  ggplot(aes("simple vs cori", estimation_diff)) + 
    geom_violin() +
    geom_boxplot(width = 0.1, outlier.alpha = 0) +
    scale_y_log10(labels = scales::label_number(accuracy = 0.0001)) + 
    labs(y = "absolute difference in Rt estimation", x = "") +
    custom_theme()

ggarrange(panel_A, panel_B, ncol = 2, widths = c(2,1), labels = c("A","B"), common.legend = T)
```

## Using the Simple Ratio to Estimate $R_t$ with real-world COVID-19 data {.unnumbered}

We next compared $R_t$ estimates from the Simple Ratio and Cori et al for using data from the current COVID-19 pandemic. We estimated $R_t$ in four nations using a generation interval with mean of `r COVID19_params$mean_gi` and standard deviation of `r COVID19_params$sd_gi`[@ganyani_estimating_2020]. \cref{fig:real-world-data}**A** shows the results for the month of October 2020. Visual inspection shows that the two methods of estimating $R_t$ have very good concordance. Across all nations over the course of the pandemic, the median difference in estimation between the two methods was `r estimation_diff_median` (95% range of `r estimation_diff_quantile_025` to `r estimation_diff_quantile_975`) (\cref{fig:real-world-data}**B**).

# Discussion {.unnumbered}

# References {.unnumbered}

::: {#refs}
:::

\newpage

# Supplementary Material {.unnumbered}

\beginsupplement

```{=tex}
\newpage
\blandscape
```
```{r load-supp-data}
rt_estimates_meanerr <- readd("rt_estimates_meanerr")
rt_estimates_sderr <- readd("rt_estimates_sderr")

supp_error_summary.meanerr = rt_estimates_meanerr %>% filter(time > 49) %>% 
  select(matches("rt.+_mean_error")) %>% 
  rename_with(~ gsub("rt\\.(.+)_mean_error", "\\1", .x)) %>%
  transmute_all(abs) %>% 
  summarise_all(quantile, probs = c(0.025, 0.5, 0.975), na.rm = T) %>% 
  transmute_all(signif, digits = 2) %>%
  transmute_all(format, scentific = F)

supp_error_summary.sderr = rt_estimates_sderr %>% filter(time > 49) %>% 
  select(matches("rt.+_mean_error")) %>% 
  rename_with(~ gsub("rt\\.(.+)_mean_error", "\\1", .x)) %>%
  transmute_all(abs) %>% 
  summarise_all(quantile, probs = c(0.025, 0.5, 0.975), na.rm = T) %>% 
  transmute_all(signif, digits = 2) %>%
  transmute_all(format, scentific = F)
```

```{r all-simulations, fig.width = 11, fig.height = 6, fig.cap = "All Simulations"}

rt_estimates %>%
  create_estimate_plot() + 
  facet_wrap(~simulation_run) + 
  theme(strip.background = element_blank(), strip.text = element_blank())
```

\elandscape

\newpage

(ref:estimation-inaccParams-cap) \textbf{$R_t$ estimation error when the mean or standard deviation of the generation interval is mis-specified}. Similar to \textbf{Figure 1}, this figure demonstrate the distribution of absolute error when estimating $R_t$ for 100 simulated epidemics. \textbf{(A)} Distribution of errors when the mean of the generation interval is mis-specified as 2 days too long.The median absolute error for the Cori and simple method were `r supp_error_summary.meanerr$cori['50%']` (95% IPR `r supp_error_summary.meanerr$cori['2.5%']` to `r supp_error_summary.meanerr$cori['97.5%']`) and `r supp_error_summary.meanerr$simple['50%']` (95% IPR `r supp_error_summary.meanerr$simple['2.5%']` to `r supp_error_summary.meanerr$simple['97.5%']`), respectively. The average \textbf{(B)} Distribution of errors when the standard deviation is mis-specified as 3 days too long. The median absolute error for the Cori and simple method were `r supp_error_summary.sderr$cori['50%']` (95% IPR `r supp_error_summary.sderr$cori['2.5%']` to `r supp_error_summary.sderr$cori['97.5%']`) and `r supp_error_summary.sderr$simple['50%']` (95% IPR `r supp_error_summary.sderr$simple['2.5%']` to `r supp_error_summary.sderr$simple['97.5%']`), respectively.

```{r estimation-inaccParams,  fig.cap = "(ref:estimation-inaccParams-cap)"}
panel_A <- rt_estimates_meanerr %>% 
  filter(time > 49) %>%
  create_rt_estimation_error_boxplot()

panel_B <- rt_estimates_sderr %>% 
  filter(time > 49) %>%
  create_rt_estimation_error_boxplot()
            
ggarrange(panel_A, panel_B, ncol=2, align = "hv", labels=c("A", "B"), common.legend = T)
```
